---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
---

This is an example of a classic decision tree model.  Here I downloaded the test and train datasets on kaggle and ran this in R Studio. 

```{r}

library(readr)
train <- read_csv("C:/Users/....train.csv", 
     col_types = cols(Fare = col_number(), 
         Age = col_number()))
View(train)


test <- read_csv("C:/Users/...test.csv", 
     col_types = cols(Age = col_number(), 
         Fare = col_number()))
View(test)

str(train)
str(test)
```




##Data Dictionary

###Variable	Definition	Key
survival	Survival	    0 = No, 1 = Yes
pclass	Ticket class	  1 = 1st, 2 = 2nd, 3 = 3rd
sex	Sex	
Age	Age in years	
sibsp	# of siblings / spouses aboard the Titanic	
parch	# of parents / children aboard the Titanic	
ticket	Ticket number	
fare	Passenger fare	
cabin	Cabin number	
embarked	Port of Embarkation	C = Cherbourg, Q = Queenstown, S = Southampton
Variable Notes

pclass: A proxy for socio-economic status (SES)
1st = Upper
2nd = Middle
3rd = Lower

age: Age is fractional if less than 1. If the age is estimated, is it in the form of xx.5

sibsp: The dataset defines family relations in this way...
Sibling = brother, sister, stepbrother, stepsister
Spouse = husband, wife (mistresses and fianc?s were ignored)

parch: The dataset defines family relations in this way...
Parent = mother, father
Child = daughter, son, stepdaughter, stepson
Some children travelled only with a nanny, therefore parch=0 for them.

```{r}
# view the survival numbers 
table(train$Survived)

test$Survived <- NA
all_data <- rbind(train, test)

str(all_data)
head(all_data)

```

Let's get to know the data by creating a few simple visuals 
```{r}
library(ggplot2)
# Look at the distribution of sex per class 
ggplot(train,aes(x=factor(Pclass),fill=factor(Sex))) +
    geom_bar(position='dodge') +
    ggtitle("Training Set")

ggplot(all_data,aes(x=factor(Pclass),fill=factor(Sex))) +
    geom_bar(position='dodge') +
    ggtitle("All Data")

```



```{r}
# look at the distribution of sex that survived or died.  It appears that more females survived.   
# for this set we'll just look at the training data because we filled the test$Survived column with NAs to better visualize all the data together and make visual comparisons. 
ggplot(train,aes(x=factor(Survived),fill=factor(Sex))) +
    geom_bar(position='dodge') +
    ggtitle("Training Set")

#here's a look at the exact numbers
table(train$Sex, train$Survived)
```

We can view the survival distribution of children by cabin class and sex.

```{r}
# using just the training data 
children <- subset(train, Age <= 18) 
ggplot(children,aes(x=factor(Pclass),fill=factor(Sex))) +
      geom_bar(position='dodge') +
      facet_grid(.~ Survived) +
      ggtitle("Training Set")
```



```{r}
ggplot(train, aes(x=factor(Pclass), fill=factor(Sex))) +
    geom_bar(position='dodge') +
    facet_grid(.~ Survived)

# now let's view this slightly differently
posn.j <- position_jitter(0.3, 0)

ggplot(train, aes(x=factor(Pclass), y=Age, col=factor(Sex))) +
  geom_jitter(size=3, alpha=0.5, position=posn.j) +
  facet_grid(. ~ Survived)
```

Let's now look at survival in the context of the cost of ticket and gender.  It appears females that paid more for their tickets had a better chance of survival. 
```{r}
posn.j <- position_jitter(0.4, 0)

ggplot(train, aes(x=factor(Sex), y=Fare, col=factor(Sex))) +
  geom_jitter(size=3, alpha=0.5, position=posn.j) +
  facet_grid(. ~ Survived)
```



```{r}
# Create the column child, and indicate whether child or no child
train$Child <- NA
test$Child <- NA

train$Child[train$Age < 18] <- 1
train$Child[train$Age >= 18] <- 0

test$Child[test$Age < 18] <- 1
test$Child[test$Age >= 18] <- 0


#view the new column to see what's happening with the data.  There's a number of NAs where we didn't have data for age. 
train$Child
```


```{r}
#install.packages("rpart")
#install.packages('rattle')
#install.packages('rpart.plot')
#install.packages('RColorBrewer')
library(rpart)
library(rpart.plot)
library(RColorBrewer)
```


```{r}
fit <- rpart(Survived ~ Pclass + Sex + Age + SibSp + Fare + Embarked + Child, data = train, method = 'class')

```


```{r}
rpart.plot(fit)
```

```{r}

prediction <- predict(fit, test, type = "class")
```


```{r}
solution <- data.frame(PassengerId = test$PassengerId, Survived = prediction)

head(solution) # look at your survival predictions in the context of passenger ids in the test set
```


```{r}

write.csv(solution, file = "firstforest.csv", row.names = FALSE)

```









